The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-08-05 Paper/jaccardbaetica.tex
eng correct intro

Our case of study has been focused on the amphoric production from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. Amphoric stamps are used to identify the presence of different groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from two centres: 1) production centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} (Inferior and Superior) and \textit{Britannia} as consumption centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance. The analysis explores how the quantitative approach provides a useful tool for the interpretation of the economic processes. Finally, results pretend to highlight the organization of Baetican olive oil production in the Roman Empire linked to the differences observed in the archaeological evidence.

@@

The Empire developed a series of structures to support and organize this long-range trade and specialized entire regions to massively produce specific goods. An example of this process was the province of \textit{Baetica} (currently Andalusia, southern Spain) which became an important olive oil production centre during the Roman Empire. Olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase on the production which was distributed through amphorae shipped via maritime routes to all the provinces and particularly to Italy and the hundreds of garrisons that the Roman army deployed along its borders~\citep{blazquez_exportacion_1980}.

The structure and processes associated to this massive olive oil production has been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. 
However, the lack of written records has made possible to detect any indication of patterns in the olive oil market.

Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: How was amphora production organized during the Roman Empire? Can we detected economic dynamics between provinces by analysing the distribution of amphorae?

@@

Advances in the research of the Roman studies have currently led an environment with various interpretations on commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}.
%\xavi{more diverse que qué?} 
The application of different quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and reveal the dynamics of the Roman production %\xavi{en qué sentido? qué quieres decir aquí con complexity? Si no especificas queda muy genérico (rollo "el mundo es muy complejo")}.
\citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

@@

The finding of amphoric stamps in different provinces has allowed us to know the importance of the distribution of olive oil within the Empire. Despite having archaeological evidence, no consensus has been reached: How was the economical relation between production and consumption centers? Did they follow a economical pattern to distribute olive oil in those provinces?

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/general_map}
\caption{Overview of the sites analyzed in this paper. The color defines the three different regions under study (\textit{Baetica} as producer; \textit{Germania} and \textit{Britannia} as consumers) while the size of each dot highlights the number of Dressel-20 amphoric stamps found on each of these sites}

@@

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. To do this, a population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}.  A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. Stamps will be used to identify economic patterns by analysing their similarity. If workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.

>>>>

Our case of study has been focused on amphoric production from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. Amphoric stamps are used to identify the presence of different groups that might share similar stamps.

To achieve this goal, we analyse a set of stamps from two centres: 1) production centres by analysing different workshops in \textit{Baetica} province and 2) two Roman provinces such as \textit{Germania} (Inferior and Superior) and \textit{Britannia} as consumption centres. They will be used to detect a connection between the distribution of amphoric stamps and the economic structure in both centres. Here, we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance. The analysis explores how the quantitative approach provides a useful tool for the interpretation of the economic processes. Finally, results pretend to highlight the organisation of Baetican olive oil production in the Roman Empire linked to the differences observed in the archaeological evidence.

@@

The Empire developed a series of structures to support and organise this long-range trade and specialised entire regions to massively produce specific goods. An example of this process was the province of \textit{Baetica} (currently Andalusia, southern Spain) which became an important olive oil production centre during the Roman Empire. Olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase on the production which was distributed through amphorae shipped via maritime routes to all the provinces and particularly to Italy and the hundreds of garrisons that the Roman army deployed along its borders~\citep{blazquez_exportacion_1980}.

The structure and processes associated with this massive olive oil production have been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. 
However, the lack of written records has made possible to detect any indication of patterns in the olive oil market.

Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: How was amphora production organised during the Roman Empire? Can we detect economic dynamics between provinces by analysing the distribution of amphorae?

@@

Advances in the research of the Roman studies have currently led an environment with various interpretations on commercial dynamics \citep{duncan1982economy,
temin_economy_2006,
quantifyingwilson2009}.
%\xavi{more diverse que qué?} 
The application of different quantitative approaches has allowed us to improve our interpretation of the growing amount of archaeological evidence and reveal the dynamics of Roman production %\xavi{en qué sentido? qué quieres decir aquí con complexity? Si no especificas queda muy genérico (rollo "el mundo es muy complejo")}.
\citep{brughmans_roman_2016,
orengo_seeds_2016,bayesian_2018,
coto-sarmiento_identifying_2018,
rubio-campillo_ecology_2018}.

@@

The finding of amphoric stamps in different provinces has allowed us to know the importance of the distribution of olive oil within the Empire. Despite having archaeological evidence, no consensus has been reached: How was the economical relation between production and consumption centres? Did they follow an economical pattern to distribute olive oil in those provinces?

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/general_map}
\caption{Overview of the sites analysed in this paper. The color defines the three different regions under study (\textit{Baetica} as producer; \textit{Germania} and \textit{Britannia} as consumers) while the size of each dot highlights the number of Dressel-20 amphoric stamps found on each of these sites}

@@

This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. To do this, a population approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}.  A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here, we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potentially spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. Stamps will be used to identify economic patterns by analysing their similarity. If workshops and provinces share stamps with similar traits, then we can identify connections. By contrast, if we do not detect similar stamps between workshops, then workshops worked independently.
